Elephant Seal Age JAE Intraspecific Variation

### 2024-05-22 paper/paperv3.qmd
p(breed) split regression in 2

```{r}
#| label: setup
#| include: false  
library(ggthemes)
library(cowplot)
library(lme4)
library(lmerTest) # Loading this library will automatically append p-values to lmer models estimated using something called Satterthwaite's method
library(merTools)
library(tidyverse)
library(ggeffects)
library(patchwork)
library(segmented)
source(here::here("R/format.R"))

sealdat <- read_csv(here::here("data/raw/128L pull 2023_12_05.csv"), 
                    show_col_types = FALSE) %>% 
  mutate(observed = if_else(observed == "B", "Breeder", "Non-breeder"), 
         observed_int = if_else(observed == "Breeder", 1, 0), 
         pup_survived = pupseeneveragain > 0) %>% 
  filter(age > 3, 
         year > 1987) %>% 
  mutate(animalID = factor(animalID),
         year_fct = factor(year),
         age10 = age / 10)

@@

We predicted that breeding success would decline with age. We found that breeding success increases to prime age, then decreases at older ages (@fig-age_dist).

>>>>

```{r}
#| label: setup
#| include: false  
library(ggthemes)
library(cowplot)
library(lme4)
library(lmerTest) # Loading this library will automatically append p-values to lmer models estimated using something called Satterthwaite's method
library(merTools)
library(tidyverse)
library(ggeffects)
library(ggridges)
library(patchwork)
source(here::here("R/format.R"))

age_senesce <- 12

sealdat <- read_csv(here::here("data/raw/128L pull 2023_12_05.csv"), 
                    show_col_types = FALSE) %>% 
  mutate(observed = if_else(observed == "B", "Breeder", "Non-breeder"), 
         observed_int = if_else(observed == "Breeder", 1, 0), 
         pup_survived = pupseeneveragain > 0) %>% 
  filter(age > 3, 
         year > 1987) %>% 
  mutate(animalID = factor(animalID),
         year_fct = factor(year),
         age10 = age / 10) %>% 
  group_by(animalID) %>% 
  mutate(longevity = ifelse(max(year) < 2020,
                            max(age),
                            NA),
         longevity10 = longevity / 10) %>% 
  ungroup()

@@

~~We predicted that breeding success would decline with age. We found that breeding success increases to prime age, then decreases at older ages (@fig-age_dist).~~

```{r}
#| label: breeding-mod 
#| fig-width: 9
#| fig-height: 6

breed_age_mod1 <- glmer(
    observed_int ~ age10 + (1 | animalID) + (1 | year_fct),
    data = filter(sealdat, age < age_senesce),
    family = "binomial",
    control = glmerControl(optimizer="bobyqa")
)
breed_age_mod2 <- glmer(
    observed_int ~ age10 + (1 | animalID) + (1 | year_fct),
    data = filter(sealdat, age >= age_senesce),
    family = "binomial",
    control = glmerControl(optimizer="bobyqa")
)

breed_pred <- rbind(
  ggpredict(
    breed_age_mod1,
    terms = c("age10 [all]"),
    type = "fixed",
    interval = "confidence"
  ) %>% 
    as_tibble() %>% 
    mutate(age = x * 10,
           age_class = "young"),
  ggpredict(
    breed_age_mod2,
    terms = c("age10 [all]"),
    type = "fixed",
    interval = "confidence"
  ) %>% 
    as_tibble() %>% 
    mutate(age = x * 10,
           age_class = "old")
) %>% 
  filter(age <= 20)

get_ranefs <- function(model, r, reduce = TRUE) {
  result <- ranef(model)[[r]] %>% 
    as_tibble(rownames = r) %>% 
    rename(ranef = 2) %>% 
    arrange(ranef)

  if (reduce)
    slice(result, round(seq(1, nrow(result), length.out = 5)))
  else 
    result
}

year_ranef1 <- get_ranefs(breed_age_mod1, "year_fct")
year_ranef2 <- get_ranefs(breed_age_mod2, "year_fct")
indiv_ranef1 <- get_ranefs(breed_age_mod1, "animalID")
indiv_ranef2 <- get_ranefs(breed_age_mod2, "animalID")

get_ranef_pred <- function(model, r, keep) {
  ggpredict(model,
            terms = c("age10 [all]", r), 
                             type = "random",
                             interval = "confidence") %>% 
  as_tibble() %>% 
  mutate(age = x * 10,
         "{r}" := group) %>% 
  semi_join(keep, by = r) %>% 
  filter(age <= 20)
}

breed_pred_year <- rbind(
  get_ranef_pred(breed_age_mod1, "year_fct", year_ranef1) %>% 
    mutate(age_class = "young"),
  get_ranef_pred(breed_age_mod2, "year_fct", year_ranef2) %>% 
    mutate(age_class = "old")
) %>% 
  filter(age <= 20)

breed_pred_indiv <- rbind(
  get_ranef_pred(breed_age_mod1, "animalID", indiv_ranef1) %>% 
    mutate(age_class = "young"),
  get_ranef_pred(breed_age_mod2, "animalID", indiv_ranef2) %>% 
    mutate(age_class = "old")
) %>% 
  filter(age <= 20)

breed_nonbreed <- sealdat %>% 
  group_by(age) %>% 
  summarize(n_observed = n(), 
            perc_breed = sum(observed_int) / n_observed, 
            se = (perc_breed * (1 - perc_breed) / n_observed)^0.5) %>% 
  filter(age <= 20)

ggplot(breed_pred_year, aes(age)) +
  # Prediction across years (fixed effect only)
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, 
                  group = age_class),
              breed_pred,
              fill = "grey", alpha = 0.5) +
  geom_line(aes(y = predicted, color = age_class), 
            breed_pred,
            linewidth = 1.5) +
  # Predictions for specific years (random effect)
  geom_line(aes(y = predicted, group = interaction(year_fct, age_class)),
            linewidth = 0.75, alpha = 0.5) +
  # Raw data
  geom_pointrange(aes(y = perc_breed, 
                      ymin = perc_breed - se, 
                      ymax = perc_breed + se),
                  breed_nonbreed) + 
  # Year labels
  geom_text(aes(y = predicted, label = year_fct),
            filter(breed_pred_year, age == 20),
            x = 20.2,
            hjust = 0,
            size = 3) +
  geom_text(aes(y = predicted, label = year_fct),
            filter(breed_pred_year, age == 4),
            x = 3.8,
            hjust = 1,
            size = 3) +
  scale_x_continuous("Age", limits = c(3.5, 20.5)) +
  scale_y_continuous("Breeding %", 
                     limits = c(NA, 1), 
                     labels = scales::percent) +
  scale_color_manual(values = c("cornflowerblue", "firebrick")) +
  theme_classic() +
  theme(legend.position = "none")

```{r}
#| label: age_breeding_b4after

senescence_age <- 12

breeding_mod_b4 <- glmer(
  observed2 ~ age10 + (1 | animalID) + (1 | year_fct),
  data = filter(all_years, age <= senescence_age), 
  family = "binomial",
  control = glmerControl(optimizer="bobyqa")
)

breeding_mod_after <- glmer(
  observed2 ~ age10 + (1 | animalID) + (1 | year_fct),
  data = filter(all_years, age > senescence_age), 
  family = "binomial",
  control = glmerControl(optimizer="bobyqa")
)

summary(breeding_mod_b4)
summary(breeding_mod_after)

```{r}
#| label: fig-age_breeding
#| fig-cap: "A: Age structure of northern elephant seals at Año Nuevo. B: Percentage of each age class that were observed breeding."
#| fig-width: 9

age_dist <- all_years %>% 
  filter(age < 22) %>% 
  group_by(age) %>% 
  count(age) %>% 
  ggplot(aes(x = age, y = n)) + 
  geom_col(fill = "#829BD1") + 
  labs(x = "Age (Years)", y = "Number of individuals") + 
  theme_few() + 
  theme(axis.title = element_text(size = 13))

breed_pred <- rbind(
  ggpredict(breeding_mod_b4,
            terms = "age10 [all]") %>% 
    as_tibble(),
  ggpredict(breeding_mod_after,
            terms = "age10 [all]") %>% 
    as_tibble()
) %>% 
    mutate(age = x * 10,
           age_class = ifelse(age <= senescence_age,
                              "Before",
                              "After"))

breed_nonbreed <- all_years %>% 
  group_by(age) %>% 
  summarize(n_observed = sum(!is.na(observed2)), 
            perc_breed = sum(observed2 == "Breeder") / n_observed, 
            se = (perc_breed * (1 - perc_breed) / n_observed)^0.5) %>% 
  filter(age < 22) %>%
  ggplot(aes(x = age, y = perc_breed)) + 
  geom_ribbon(aes(age, ymin = conf.low, ymax = conf.high, group = age_class),
              breed_pred,
              inherit.aes = FALSE,
              fill = "#829BD1",
              alpha = 0.25) +
  geom_line(aes(age, predicted, group = age_class),
            breed_pred,
            inherit.aes = FALSE,
            color = "#829BD1",
            linewidth = 1) +
  geom_point() + 
  geom_errorbar(aes(ymin = perc_breed - se, ymax = perc_breed + se), 
                width = 0) +
  scale_y_continuous(labels = scales::percent, limits = c(0.4, 1)) +
  expand_limits(x = 23) +
  labs(x = "Maternal Age (Years)", y = "% Breeders (Mean ± SE)") + 
  theme_few() + 
  theme(axis.title = element_text(size = 13))
